The Precambrian paleogeography of Laurentia

### 2020-10-15 Manuscript/Laurentia_Paleogeo_Manuscript.tex
edits/additions to Baltica connection text

Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period at the end of the Neoproterozoic Era where there are inconsistencies between poles of similar age (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 583 and 565 Ma (Fig. \ref{fig:Laurentia_paleolatitude}). This conflicting record is a longstanding problem and has led to the presentation of both high-latitude and low-latitude Laurentia paleogeographic reconstructions at the time (e.g. \citealp{Pisarevsky2001a,Li2008a}). One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran where rapid rotation of the entire silicate Earth influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}. Paleodirectional data from single feldspar crystals from the Sept-\^Iles layered intrusion led \cite{Bono2015a} to interpret the lower inclination (and therefore lower latitude) direction from the intrusion (the one included as the ca. 565 Ma Sept-\^Iles pole in Table 2; \citealp{Tanczyk1987a}) as the primary thermal remanent magnetization.  \cite{Bono2015a} interpret steeper directions also recovered from the intrusives as the result of remagnetization. They suggest that other steep magnetizations from Ediacaran Laurentia plutonic rocks, such as that observed in the ca. 583 Ma Baie des Mountons complex (the A group of \cite{McCausland2011a} in Table 2), are also the result of remagnetization. The lower inclination Baie des Mountons complex B Group directions result in a pole that is indistinguishable from the lower inclination  Sept-\^Iles intrusives pole. Another possibility discussed in the literature is that the lack of congruency between poles in this time interval is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}. Data from the ca. 585 Ma Grenville dyke swarm of Laurentia interpreted as primary reveal $\sim$90\textdegree\ differences in direction within dikes dated within 2.5 $\pm$ 0.9 million years of one another \citep{Halls2015a}. The rates of $>$26\textdegree /Myr ($>$288 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions. Estimates of magnetic paleointensity in these Grenville dikes are anomalously weak which could support a deviation from stable axial dipolar geomagnetic field behavior at the time \citep{Thallner2020a}.  Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the compiled record for Laurentia (Fig. \ref{fig:Laurentia_paleolatitude}).

@@

Based on correlation of Archean provinces and Paleoproterozoic orogenic belts, \cite{Gower1990a} reconstructed Baltica to Laurentia in a position known as the NENA (northern Europe and North America) configuration. This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}. In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogen of Baltica, where accretionary orogenesis was active ca. 1.66 to 1.52 Ga \citep{Bergstrom2020a}, is a continuation of the Mazatzal-Labradorian orogenic belts of Laurentia\citep{Karlstrom2001a}. Baltica and Laurentia as conjoined cratons with a shared active margin features as a major component of Paleoproterozoic to Mesoproterozoic paleogeographic reconstructions \citep{Evans2011a, Zhang2012a, Elming2021a}. Rotating Baltica into the NENA connection position results in matched paleomagnetic pole pairs between ca. 1.75 to 1.26 Ga \citep{Buchan2000a, Evans2008a} that could be extended to ca. 1.12 Ga if a virtual geomagnetic pole from the Salla dike is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}. These data constrain the NENA connection to have been maintained until at least 1.26 Ga and perhaps until 1.12 Ga \citep{Salminen2021b}.

Many paleogeographic models consider there to still have been close proximity between Baltica and Laurentia in the latest Mesoproterozoic into the Neoproterozoic. This continued connection has been hypothesized based on correlation of the Grenvillian orogeny with the Sveconorwegian orogeny of southwestern Baltica\citep{Gower1990b} as well as proposed similarities between the apparent polar wander path of Laurentia's Grenville loop and Baltica's Sveconorwegian loop that led reconstructions to be based on the alignment of these paths and the resulting position of Baltica relative to Laurentia \citep{Piper1980a,Pisarevsky2003a}. However, as geochronology has improved the ages of poles in the Sveconorwegian loop are now constrained to be younger than the Grenville loop poles falling in a gap in Laurentia's pole record \citep{Evans2015a,Fairchild2017a} which makes it more challenging to test proposed post-NENA configurations between the cratons.

In terms of the orogenic timing, the main phase of the Sveconorwegian orogeny was from ca. 1.05 to 0.98 Ga corresponds temporally with the Grenvillian orogeny \citep{Stephens2020a}. However, it is currently an active debate in the literature whether the Sveconorwegian orogen is the result of collisional or accretionary orogenesis \citep{Stephens2020a}. \cite{Slagstad2019a} favor an accretionary orogeny and by contrasting this setting with the collisional orogenesis of the Grenvillian orogen argue that there should not be a link between Laurentia and Baltica ca. 1.0 Ga. However, in a recent critical review of the constraints, the high pressure nature of metamorphism and the cratonward propagation of orogenesis (in contrast to older accretionary orogenesis) led \cite{Stephens2020a} to favor the interpretation that the Sveconorwegian orogen is a record of prolonged continent-continent collision initiating ca. 1.06 Ga.

Another basis for a continued connection between

A clockwise rotation of Baltica relative to Laurentia from a Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador, Canada is implemented in many paleogeographic models (e.g. \citealp{Evans2009a})...

% orogen in Labrador         as advocated by \cite{Gower1990b} 
%
%Tonian period (ca. 975 to 915 Ma; \citealp{McClelland2019a}) intrusive rocks in Svalbard and East Greenland are interpreted to be related to subduction and accretionary orogenesis that has been termed the Valhalla orogeny \citep{Cawood2010a}.
%
%The NENA connection 
%
% -- a constraint that is honored in . This reconstruction is brings paleomagnetic poles
%
%compatible with existing paleomagnetic constraints from ca. 1750 to 1270 Ma \cite{Evans2008a} and these conjoined cratons feature as a major component of the hypothesized Nuna supercontinent (\citealp{Evans2011a, Zhang2012a}; .

>>>>

Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period at the end of the Neoproterozoic Era where there are inconsistencies between poles of similar age (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 583 and 565 Ma (Fig. \ref{fig:Laurentia_paleolatitude}). This conflicting record is a longstanding problem and has led to the presentation of both high-latitude and low-latitude Laurentia paleogeographic reconstructions at the time (e.g. \citealp{Pisarevsky2001a,Li2008a}). One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran where rapid rotation of the entire silicate Earth influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}. Paleodirectional data from single feldspar crystals from the Sept-\^Iles layered intrusion led \cite{Bono2015a} to interpret the lower inclination (and therefore lower latitude) direction from the intrusion (the one included as the ca. 565 Ma Sept-\^Iles pole in Table 2; \citealp{Tanczyk1987a}) as the primary thermal remanent magnetization.  \cite{Bono2015a} interpret steeper directions also recovered from the intrusives as the result of remagnetization. They suggest that other steep magnetizations from Ediacaran Laurentia plutonic rocks, such as that observed in the ca. 583 Ma Baie des Mountons complex (the A group of \cite{McCausland2011a} in Table 2), are also the result of remagnetization. The lower inclination Baie des Mountons complex B Group directions result in a pole that is indistinguishable from the lower inclination  Sept-\^Iles intrusives pole. Another possibility discussed in the literature is that the lack of congruency between poles in this time interval is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}. Data from the ca. 585 Ma Grenville dyke swarm of Laurentia interpreted as primary reveal $\sim$90\textdegree\ differences in direction within dikes dated within 2.5 $\pm$ 0.9 million years of one another \citep{Halls2015a}. The rates of $>$26\textdegree /Myr ($>$288 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions. Estimates of magnetic paleointensity in these Grenville dikes are anomalously weak \citep{Thallner2020a} as are data from coeval volcanics in Ukraine \citep{Shcherbakova2019a} which could support an anomalous deviation from stable axial dipolar geomagnetic field behavior at the time as interpreted by \cite{Halls2015a}.  Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the compiled record for Laurentia (Fig. \ref{fig:Laurentia_paleolatitude}).

@@

Based on correlation of Archean provinces and Paleoproterozoic orogenic belts, \cite{Gower1990a} reconstructed Baltica to Laurentia in a position known as the NENA (northern Europe and North America) configuration. This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}. In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogen of Baltica, where accretionary orogenesis was active ca. 1.66 to 1.52 Ga \citep{Bergstrom2020a}, is a continuation of the Mazatzal-Labradorian orogenic belts of Laurentia\citep{Karlstrom2001a}. Baltica and Laurentia as conjoined cratons with a shared active margin features as a major component of Paleoproterozoic to Mesoproterozoic paleogeographic reconstructions \citep{Evans2011a, Zhang2012a, Elming2021a}. Rotating Baltica into the NENA connection position results in matched paleomagnetic pole pairs between ca. 1.75 to 1.26 Ga \citep{Buchan2000a, Evans2008a} that could be extended to ca. 1.12 Ga if a virtual geomagnetic pole from the Salla dike of northeastern Finland is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}. These data constrain the NENA connection to have been maintained until at least 1.26 Ga and perhaps until 1.12 Ga \citep{Salminen2021b}.

Many paleogeographic models consider there to still have been close proximity between Baltica and Laurentia in the latest Mesoproterozoic into the Neoproterozoic. This continued connection has been hypothesized based on correlation of the Grenvillian orogeny with the Sveconorwegian orogeny of southwestern Baltica\citep{Gower1990b} as well as proposed similarities between the apparent polar wander path of Laurentia's Grenville loop and Baltica's Sveconorwegian loop that led reconstructions to be based on the alignment of these paths and the resulting position of Baltica relative to Laurentia \citep{Piper1980a,Pisarevsky2003a}. However, as geochronology has improved, the ages of poles in the Sveconorwegian loop are now constrained to be younger than the Grenville loop poles falling in a gap in Laurentia's pole record \citep{Evans2015a,Fairchild2017a} which makes it more challenging to test proposed post-NENA configurations between the cratons.

In terms of the orogenic timing, the main phase of the Sveconorwegian orogeny from ca. 1.05 to 0.98 Ga corresponds temporally with the Grenvillian orogeny \citep{Stephens2020a}. It is currently an active debate in the literature whether the Sveconorwegian orogen is the result of collisional or accretionary orogenesis \citep{Stephens2020a}. \cite{Slagstad2019a} favor an accretionary orogeny and by contrasting this setting with the collisional orogenesis of the Grenvillian orogen argue that there should not be a link between Laurentia and Baltica ca. 1.0 Ga. However, in a recent critical review of the constraints, the high pressure nature of metamorphism and the cratonward propagation of orogenesis (in contrast to older accretionary orogenesis) led \cite{Stephens2020a} to favor the interpretation that the Sveconorwegian orogen is a record of prolonged continent-continent collision initiating ca. 1.06 Ga.

Intrusive granitoids and extrusive calc-alkaline volcanics with early Tonian Period ages between 975 to 915 Ma are present in East Greenland and related terranes including East Svalbard \citep{McClelland2019a}. These intrusive and volcanic rocks are interpreted to have formed within a magmatic arc or in a syn-collisional setting \citep{Johansson1999a} that has been hypothesized to be the result of subduction and accretionary orogenesis termed the Valhalla orogeny \citep{Cawood2010a}. The need for the East Greenland margin to have been active in the earliest Neoproterozoic has motivated a tectonic model where Baltica rifted off East Greenland in the late Mesoproterozoic while staying proximal to Laurentia via a clockwise rotation \citep{Cawood2010a}. Such a clockwise rotation of Baltica relative to Laurentia from a Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador, Canada is implemented in many paleogeographic models (e.g. \citealp{Evans2009a} and as shown in Fig. \ref{fig:Grenville_reconstructions}). A challenge with this model is it is unclear what was the conjugate lithosphere that rifted from East Greenland leading to thick sedimentary successions interpreted to have been deposited on a thermally-subsiding passive margin from ca. 850 Ma into the Cryogenian (Fig. \ref{fig:tectonic_history}; \citealp{Maloof2006a}). \cite{Malone2014a} interpreted this Neoproterozoic basin development along the East Greenland margin to have been the result of extension in a back-arc setting although the paucity of tuffs within successions such as the Akademikerbreen Group of East Svalbard is difficult to reconcile with such a setting. The separation of Baltica from Laurentia is hypothesized to have occurred ca. 600 Ma associated with the opening of the Iapetus Ocean (Fig. \ref{fig:tectonic_history}; \citep{Cawood2001a}).
